NumLin: Linear Types for Numerical Libraries

### 2018-10-25 write-up/paper/implementation.tex
Start desugaring

\lang\ transpiles to OCaml and its implementation follows the structure of a
typical domain-specific language compiler. Although \lang's current
implementation is not as embedded DSL, its the general design is simple enough
to adapt to being so and also to target other languages.

Alongside the transpiler, a `Read-Check-Translate' loop, benchmarking program
and a test suite are included in executable artifacts accompanying this paper.

@@

A very pleasant way to use \lang\ is to have the build system generate code at
\emph{compile-time} and then have it be used by other modules like normal OCaml
functions. This makes it possible and even easy to use \lang\ alongside
existing OCaml libraries; in fact, this is exactly how the benchmarking program
and test-suite use code written in \lang.

@@

type 'a bang =
  Many of 'a
  [@@unboxed]
        \end{minted}
    \end{minipage}
    \begin{minipage}{.3\textwidth}
        \begin{align*}
            [\![ f\!c ]\!] &= \texttt{'fc} \\
            [\![ \textbf{Z} ]\!] &= \texttt{z}\\
            [\![ \textbf{S} \, f ]\!] &= [\![ f ]\!]\, \texttt{s}\\
            [\![ \textbf{unit} ]\!] &= \texttt{unit}\\
            [\![ \textbf{bool} ]\!] &= \texttt{bool}\\
            [\![ \textbf{int} ]\!] &= \texttt{int}\\
            [\![ \textbf{elt} ]\!] &= \texttt{float}\\
            [\![ f\, \textbf{arr} ]\!] &= [\![ f ]\!]\, \texttt{arr}\\
            [\![ f\, \textbf{mat} ]\!] &= [\![ f ]\!]\, \texttt{mat}\\
            [\![ \textbf{!} \, t ]\!] &= [\![ t ]\!]\, \texttt{bang}\\
            [\![ \forall f\!c.\, t ]\!] &= [\![ t ]\!]\\
            [\![ t \otimes t' ]\!] &= [\![ t ]\!] \texttt{*} [\![ t' ]\!]\\
            [\![ t \multimap t' ]\!] &= [\![ t ]\!] \rightarrow [\![ t' ]\!]
        \end{align*}
    \end{minipage}
    \caption{\lang's type grammar (left) and its embedding into OCaml
        (right).}\label{fig:type_grammar}
\end{figure}

Straightforward enough.

A few examples?

>>>>

\lang\ transpiles to OCaml and its implementation follows the structure of a
typical domain-specific language (DSL) compiler. Although \lang's current
implementation is not as embedded DSL, its the general design is simple enough
to adapt to being so and also to target other languages.

Alongside the transpiler, a `Read-Check-Translate' loop, benchmarking program
and a test suite are included in the artifacts accompanying this paper.

@@

A very pleasant way to use \lang\ is to have the build system generate code at
\emph{compile-time} and then have the generated code be used by other modules
like normal OCaml functions. This makes it possible and even easy to use \lang\
alongside existing OCaml libraries; in fact, this is exactly how the
benchmarking program and test-suite use code written in \lang.

Desugaring is conventional.

\begin{align*}
    \mathbf{fun}\ (x : t)\ 'r\ (y : {'r}\ \mathbf{mat}) \rightarrow e
    & \Rightarrow
    \mathbf{fun}\ (x : t) \rightarrow \mathbf{fun}\ {'r} \rightarrow
    \mathbf{fun}\ (y : {'r}\ \mathbf{mat}) \rightarrow e
\\
    x[e]
    & \Rightarrow
    \mathbf{get}\ \_\ x\ (e) \qquad \textrm{(similarly for matrices)}
\\
    x[e_1] := e_2
    & \Rightarrow
    \mathbf{set}\ x\ (e_1)\ (e_2) \qquad \textrm{(similarly for matrices)}
\\
    \mathbf{let}\ !x = e_1\ \mathbf{in}\ e_2
    & \Rightarrow
    \mathbf{let\ Many}\ x = e_1\ \mathbf{in\ }
    \mathbf{let\ Many}\ x = \mathbf{Many}\ (\mathbf{Many}\ x)\ \mathbf{in}\ e_2
\\
    \mathbf{let}\ f\ {\langle args \rangle} = e_1\ \mathbf{in}\ e_2
    & \Rightarrow
    \mathbf{let}\ f = \mathbf{fun}\ {\langle args \rangle} \rightarrow e_1\
    \mathbf{in}\ e_2
\\
    \mathbf{let}\ !f\ {\langle args \rangle} = e_1\ \mathbf{in}\ e_2
    & \Rightarrow
    \mathbf{let\ Many}\ f = \mathbf{Many}\ (\mathbf{fun}\ {\langle args \rangle}
    \rightarrow e_1)\ \mathbf{in}\ e_2
\\
    \mathbf{let\ rec}\ f\ (x : t)\ {\langle args \rangle} : {t'} = e_1\ \mathbf{in}\ e_2
    & \Rightarrow
    \mathbf{let\ Many}\ f = \mathbf{Many}\ (\mathbf{fix}\ (f, x : t,
    \mathbf{fun}\ {\langle args \rangle} \rightarrow e_1 : {t'} ))\ \mathbf{in}\ e_2
\end{align*}

@@

type 'a bang =
  Many of 'a
  [@@unboxed]
        \end{minted}
    \end{minipage}
    \begin{minipage}{.3\textwidth}
        \begin{align*}
            [\![ f\!c ]\!] &= \texttt{'fc} \\
            [\![ \textbf{Z} ]\!] &= \texttt{z}\\
            [\![ \textbf{S} \, f ]\!] &= [\![ f ]\!]\, \texttt{s}\\
            [\![ \textbf{unit} ]\!] &= \texttt{unit}\\
            [\![ \textbf{bool} ]\!] &= \texttt{bool}\\
            [\![ \textbf{int} ]\!] &= \texttt{int}\\
            [\![ \textbf{elt} ]\!] &= \texttt{float}\\
            [\![ f\, \textbf{arr} ]\!] &= [\![ f ]\!]\, \texttt{arr}\\
            [\![ f\, \textbf{mat} ]\!] &= [\![ f ]\!]\, \texttt{mat}\\
            [\![ \textbf{!} \, t ]\!] &= [\![ t ]\!]\, \texttt{bang}\\
            [\![ \forall f\!c.\, t ]\!] &= [\![ t ]\!]\\
            [\![ t \otimes t' ]\!] &= [\![ t ]\!] \texttt{*} [\![ t' ]\!]\\
            [\![ t \multimap t' ]\!] &= [\![ t ]\!] \rightarrow [\![ t' ]\!]
        \end{align*}
    \end{minipage}
    \caption{\lang's type grammar (left) and its embedding into OCaml
        (right).}\label{fig:type_grammar}
\end{figure}
A few examples?

### 2018-10-25 write-up/paper/lang_and_examples.tex
Start desugaring

    \item \texttt{potrs \_ chol\_r data\_2} uses a pre-computed Cholesky decomposition
        to multiply \texttt{data\_2} by \texttt{r\_2} $= (R + H \Sigma H^T)^{-1}$

@@

With expressions as complex as those in Figure~\ref{fig:kalman_eqns}, it can
get very difficult to ensure memory is not being re-used incorrectly/enough
whilst also enusring the aliasing assumptions for the BLAS/LAPACK routine calls
are guaranteed. However, using \lang\ makes incorrect aliasing and memory
re-use impossible. Additionally, \lang\ makes it possible to spot where memory
is not being re-used enough \emph{precisely because} of its explicit handling
of allocations, aliases and frees.

>>>>

    \item \texttt{potrs \_ chol\_r data\_2} uses a pre-computed Cholesky decomposition
        to multiply \texttt{data\_2} $ = H \mu - \textrm{data}$ by
        \texttt{r\_2} $= (R + H \Sigma H^T)^{-1}$

@@

With expressions as complex as those in Figure~\ref{fig:kalman_eqns}, it can
get very difficult to ensure memory is not being re-used incorrectly/enough
whilst also enusring the aliasing assumptions for the BLAS/LAPACK routine calls
are guaranteed. However, using \lang\ makes incorrect aliasing and memory
re-use impossible. Additionally, \lang\ makes it possible to spot where memory
is not being re-used enough \emph{precisely because} of its explicit
allocations, aliases and frees.
